---
title: "Course5_Assign1"
author: "Tahrul Amin"
date: "August 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Calling required libraries/R packages
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(lubridate)
library(ggplot2)
library(lattice)
library(markdown)
```
### Part1: Loading and processing the data 
```{r message=FALSE, warning=FALSE}
setwd("C:/coursera_project/cr5wk2/FINAL")
data <- read.csv("activity.csv", header =TRUE,sep = ',', colClasses =c("numeric","character","integer"))
data$date <- ymd(data$date) #using lubridate package
```
##----------------------------------------------------------------------------------------------------------
### Part2: What is mean total number of steps taken per day?
Histogram of the total number of steps taken each day (excluding 'NA' values)
```{r message=FALSE, warning=FALSE}
steps <- data %>%filter(!is.na(steps)) %>%group_by(date) %>%summarize(steps=sum(steps)) %>%print
```
Including Plots
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(steps, aes(x = steps)) +
  geom_histogram(fill = "blue", binwidth = 2000) +
  labs(title = "Histogram of Steps per day", x = "Steps per day", y = "Frequency")
```
  
Mean and median number of steps taken each day
```{r}
mean(steps$steps)
median(steps$steps)
```
##--------------------------------------------------------------------------------------------------------
###Part3: What is the average daily activity pattern?
Time series plot of the average number of steps taken  
Compute the means of steps accross all days for each interval
```{r, echo=FALSE, message=FALSE, warning=FALSE}
mean_interval <- aggregate(data$steps,by=list(data$interval), FUN=mean,na.rm=TRUE)
names(mean_interval) <- c("interval","mean")
```
  
  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
plot(mean_interval$interval,mean_interval$mean,type="l", 
     xlab="5-min Interval", 
     ylab="Average number of steps taken", 
     main="Time series plot of the average number of steps taken (without NA)" )
```
    
The 5-minute interval that, on average, contains the maximum number of steps
```{r}
max_index <- which(mean_interval$mean == max(mean_interval$mean))
max_interval <- mean_interval[max_index, 1]
```
  
max_interval value is 835
  
##------------------------------------------------------------------------------------------------------------
  
###Part4:	Imputing missing values
Code to describe and show a strategy for imputing missing data, Histogram
  

```{r message=FALSE, warning=FALSE}
#Reading fresh data
setwd("C:/coursera_project/cr5wk2/FINAL")
data <- read.csv("activity.csv", header =TRUE,sep = ',', colClasses = c("numeric","character","integer"))
data$date <- ymd(data$date)
NA_count <- sum(is.na(data$steps))
# NA count is 2304
na_index <- which(is.na(data$steps))

#keeping very simple, replacing NA by avergae step and put it in one dim array/vector
simple_mean_to_update_NA <- rep(mean(data$steps, na.rm=TRUE), times=NA_count)
#Update dataset with missing values
data[na_index, "steps"] <- simple_mean_to_update_NA
head(data)
```
  
Histogram of the total number of steps taken each day (with replaced 'NA' values)
```{r message=FALSE, warning=FALSE}
steps <- data %>%filter(!is.na(steps)) %>%group_by(date) %>%summarize(steps = sum(steps)) %>%print
ggplot(steps, aes(x = steps)) +
  geom_histogram(fill = "blue", binwidth = 2000) +
  labs(title = "Histogram of Steps per day", x = "Steps per day", y = "Frequency")

```
  
mean and median total number of steps taken per day
```{r}
mean(steps$steps)
median(steps$steps)
```
  
The mean has not differed from the estimates from the first part of the assignment. The median differed a bit.
the impact of imputing missing data on the estimates of the total daily number of steps is low
  
##-------------------------------------------------------------------------------------------------------------
  
##Part-5: Are there differences in activity patterns between weekdays and weekends?
Using the updated dataset with the filled-in missing values
find weekdays from the date attribute
```{r}
updated_dataset <- data.frame(date=data$date, weekday=tolower(weekdays(data$date)), steps=data$steps, interval=data$interval)
# Compute new factor variable 'daytype' (two levels - weekend,weekday) and add this new column using cbind
updated_dataset <- cbind(updated_dataset, daytype=ifelse(updated_dataset$weekday == "saturday" | updated_dataset$weekday == "sunday", "weekend","weekday"))
head(updated_dataset)
# Create the final data frame
final_dataset <- data.frame(date=updated_dataset$date, weekday=updated_dataset$weekday, daytype=updated_dataset$daytype, 
                       interval=updated_dataset$interval,
                       steps=data$steps)


```

```{r}
# Compute the average number of steps taken, averaged across all daytype variable
second_mean_data <- aggregate(final_dataset$steps, 
                       by=list(final_dataset$daytype, 
                               final_dataset$weekday, final_dataset$interval), mean)

# Rename the attributes
names(second_mean_data) <- c("daytype", "weekday", "interval", "mean")
head(second_mean_data)
# Compute the time serie plot
xyplot(mean ~ interval | daytype, second_mean_data, type="l",lwd=1,xlab="Interval",ylab="Number of steps",layout=c(1,2))

```




